Hydrous melting and partitioning in and above the mantle transition zone: insights from water-rich MgO-SiO$_2$-H$_2$O experiments

### 2015-07-09 MSH.tex
Update

\title{Hydrous melting and partitioning in the deep upper mantle: insights from experiments in the system MgO-SiO$_2$-H$_2$O}
%\tnotetext[mytitlenote]{Fully documented templates are available in the elsarticle package on \href{http://www.ctan.org/tex-archive/macros/latex/contrib/elsarticle}{CTAN}.}

@@

In this study, we conducted high pressure multianvil experiments at 13 GPa between 1200 and 1900 $^{\circ}$C to investigate the liquidus in the system MgO-SiO$_2$-H$_2$O. Water-rich starting compositions were obtained by using platinic acid (H$_2$Pt(OH)$_6$) as a novel water source. As MgO:SiO$_2$ ratios decrease, the lquidus curve develops an increasingly pronounced concave-down topology. At slab temperatures, the forsterite- and enstatite-water liquidii have similar water contents. Preliminary experiments on the SiO$_2$-H$_2$O binary reveal a pronounced melting point depression up to about 20 mol\% H$_2$O. This melting point depression exceeds estimates from published models, and requires either significant non-ideality between species in the melt, or distinct differences in the ease of protonation of bridging oxygens. We discuss the implications of this work on the migration of deep-seated slab fluids, and the partitioning of water between nominally anhydrous phases and melts.
\end{abstract}

@@

\section{Introduction}
Subduction zones are primary vectors for water transport into the deep Earth. This water is progressively lost as fluids and then melts as the surrounding mantle warms the slab. The most prominent surface expression of this melting is the presence of arc volcanoes about 100 km above the slab surface, but hydrous minerals within the crust should allow some water to be transported to 250 km, and peridotite hydration should allow water to penetrate into the lowermost mantle transition zone, providing it remains low enough in temperature. A startling confirmation of such deep water is the discovery of water-saturated ringwoodite in a kimberlitic diamond from Juina, Brazil.

Several key questions present themselves. When water-rich melts are created in subduction zones, what are their compositions? How do their compositions and resulting physical properties (density, viscosity, melt-solid dihedral angle) change as they migrate? Where will they migrate? Their migration paths will have major implications for figuring out the water budget of the Earth - if they are trapped close to or within the slab, hydrating the whole of the mantle transition zone will be much less likely.

There is another reason to be interested in the compositions of hydrous melts. The thermal stability of dense hydrous magnesium silicate (DHMS) phases and the partitioning of water between silicate melt and nominally anhydrous minerals (NAMS) control water transport into the deep mantle. At shallow depths, the fluids released during dehydration of these phases contain only a few percent of dissolved solutes. As such, it is reasonable to simplify thermodynamic calculations by approximating the fluid product of dehydration as pure water. As pressure increases, several processes ensure that this is no longer the case. Firstly, the second critical endpoint of peridotitic and eclogitic systems occur at <10 GPa, after which there is no longer a distinction between melt and fluid, and melt compositions smoothly migrate towards lower water contents with increasing temperature. Secondly, the thermal stability of DHMS phases increases. As a result, the final breakdown of DHMS phases at mantle transition zone pressures occurs when MgO-SiO$_2$-H$_2$O melts have H$_2$O activities significantly lower than one, which will also strongly influence the water content of NAMS. Accurate thermodynamic estimates of water storage capacities in mantle rocks thus require models for both the silicate phases and the coexisting melts.

% There has been some controversy over whether slabs transport water beyond about 300 km depth. Certainly most of the ca. 7 km thick mafic crust is probably too warm to sustain hydrous minerals. However, hydrated ultramafics can potentially carry water into the lower mantle. Supporting evidence for an ultradeep part of the water cycle exists in the form of hydrous ringwoodite in diamond, which apparently formed under cold conditions.

% The compositions of melts formed by the breakdown of hydrous phases is important for a full understanding of the deep water system. The thermodynamics of melting in water-rich systems cannot be determined without good estimates of melt compositions. Furthermore, the thermal stability of hydrous phases cannot be predicted from their properties without also having an activity-composition model for the melt phase. More generally, the density of hydrous melts will be highly composition dependent, impacting on their ability to rise through the mantle. Finally, reactions between migrating melts and the surrounding solid will be controlled by the \emph{P-T-X} dependence on melt composition.

@@

In this study, we use high purity platinic acid, hexahydroxyplatinate(IV) (H$_2$Pt(OH)$_6$) as a novel water source. Platinic acid is a pale yellow compound which is stable to about 130 $^{\circ}$C \citep{Nagano2002}. It breaks down at higher temperatures, releasing four molecules of H$_2$O and forming platinum (IV) oxide PtO$_2$. Its relatively high stability, high water content and the inert nature of the breakdown product (in a system where redox reactions are absent) make it an excellent water source for high pressure hydrous melting experiments.

\section{Experimental and analytical methods}
Starting compositions were created from a mixture of high purity brucite (Mg(OH)$_2$, SiO$_2$ and platinic acid (H$_2$Pt(OH)$_6$). SiO$_2$ was dried overnight at 1000 $^{\circ}$C, while brucite was heated to 250 $^{\circ}$C overnight. Both powders were stored in a desiccating oven at 130 $^{\circ}$C. Platinic acid is hygroscopic, so was stored in a vacuum desiccator. Powders were weighed and ground dry in an agate mortar for 30 minutes, using a mask, goggles and gloves to avoid physical contact with the platinic acid. Starting compositions were stored in glass vials in a vacuum desiccator. Compositions are listed in Table \ref{table:compositions}.

@@

\section{Results}
Experimental details and results are shown in Table \ref{table:experiments}. Superliquidus runs revealed networks of dendritic crystals of brucite, forsterite, enstatite and stishovite. Below the liquidus, solid phases tended to aggregate at the cold end of the capsules. Periclase formed equant crystals often associated with quench overgrowths which appear brighter in BSE images. Forsterite formed tabular crystals about 100 $\mu$m long. Enstatite crystals were much smaller, forming ca. 5 $\mu$m equant grains. Stishovite also formed equant to blocky grains, but crystals only separated from the melts in chambers with a high water content. In addition to the difference in texture and shape between quench crystals liquidus phases, subliquidus phases typically contained many small inclusions of Pt-Rh oxides. A typical set of run products is shown in Figure \ref{fig:sem}.

\section{Discussion}
\subsection{Modelling water solubility in melts}

@@

In this study, we use a modified version of the molecular dynamics-derived anhydrous melt model of \cite{DKS2013} using melt endmembers MgO and SiO$_2$. For MgO melt, only the reference Helmholtz Free energy needed to be adjusted. For the SiO$_2$ melt, the bulk modulus of the SiO$_2$ endmember was increased to satisfy the experimental constraint that coexisting melt and coesite have similar volumes at 13--14 GPa. As a result, SiO$_2$ volumes at lower mantle conditions will be significantly higher than predicted, but this detail is not important for the present study. The volume at 1 bar is fit to experimental data. The entropy at reference conditions is adjusted by the value corresponding to the difference between the experimentally-derived and MD-calculated stishovite values. Remarkably, the value of the entropy derived in this way is essentially identical to that calculated for the experimental data of \cite{} near the coesite-stishovite-melt triple point. As such, we are confident that this value is accurate. It should be noted that the recent melting curve proposed by \cite{} (using a Simon's equation) predicts a much higher entropy of melting at 14 GPa. This is a result of downweighting the data in the study of \cite{}, which indicates a controversially high dT/dP. Nevertheless, this study is not inconsistent with higher pressure studies \citep{} if the gradient of the melting curve shallows between 30 and 70 GPa, which may be the consequence of changing Si-coordination and resultant densification in the melt.

A subregular solution model was adopted to describe the free energy between the MgO and SiO$_2$ melts. The two interaction parameters W$_{MS}$ and W$_{SM}$ 14 GPa were derived from the melt composition at the forsterite (fo)-HP clinoenstatite (hen) eutectic \citep{} and melting temperatures at the fo, hen and fo-hen cotectic compositions. The temperature dependence around 14 GPa was calculated by assuming an ideal excess entropy contribution, and then fitting to the known melting curves of forsterite and enstatite. The pressure dependencies were fit to the melting curves. It should be noted that although the excess volumes are similar to the results of \cite{DKS2013}, the excess entropies are some 50\% lower. However, they provide an excellent fit to the forsterite and orthoenstatite melting curves (see Supplementary Materials).

All of the calculations in the work were undertaken with the mineral physics toolkit Burnman \cite{}.

\subsection{Subsystems}
\subsubsection{MgO-H$_2$O}
The dehydration of brucite at high pressure was previously studied by \cite{FIYKFO2005}. They reported that the stability of brucite reaches a maximum at about 9--10 GPa and 1200 $^{\circ}$C, decomposing at 1100--1150 $^{\circ}$C at 13 GPa. We observed large brucite crystals at 1200 $^{\circ}$C and 13 GPa; the temperature difference is probably within experimental uncertainties. In the present study, the composition of the fluid coexisting with brucite and periclase is constrained to be 64--66 mol\% H$_2$O. The MgO content of the fluid increases gradually with increasing pressure, as a result of the extremely high melting temperature of periclase. The extremely high melting point of MgO means that the best experimental estimate of the melting point at high pressure (7 GPa) is derived by extrapolation from iron-bearing (Mg,Fe)O, using the compositions of coexisting melt and solid to constrain the shape of the two-phase region \citep{ZF2008}. Their melting curve is within error of that obtained from early molecular dynamics simulations \citep{CG1994}, which suggest a melting temperature of $\sim$ 5373 K at 13 GPa. In this study, we follow more recent MD simulations \citep{KKS2013, DKS2013} which predict periclase melting at a lower temperature of $\sim$ 4180 K at 13 GPa, similar to that determined by previous ab-initio studies \citep{Alfe2005}, but significantly higher than early experimental estimates \citep[$\sim$ 3100 K][]{ZB1994}.

At 13 GPa, the calculated entropy of melting of periclase is $27.5$ J/K/mol, somewhat higher than the value from previous MD simulations \cite[$\sim$22 J/K/mol][]{CG1994}. This higher value will moderate the differences in melting temperature between the two studies. Using the ideal OH$^-$ mixing model of \cite{SS1985} with no compositional dependence on the formation of OH groups, $G^{\circ}= -75000$ provides a good fit to the data if additional asymmetry is added via a binary subregular model (W$_{MgO-H_2O}$=55000 J/mol, W$_{H_2O-MgO}$=0 J/mol). This asymmetry is required to fit the activity of water at the periclase-brucite-liquid triple point to the solid thermodynamic models of \citep{HP2011} and the equation of state of water formulated by \cite{PS1995} ($0.52$). This model also has the effect of increasing the tendency for melt-fluid immiscibility at the water-rich end of the system. The second critical endpoint of the MgO-H$_2$O is probably around 13 GPa \citep{MSUP2007}, so large positive interaction enthalpies are justified in this region of the phase diagram.

@@

\clearpage
\subsubsection{SiO$_2$-H$_2$O}
The 13 GPa SiO$_2$-H$_2$O binary phase diagram is shown in Figure \ref{fig:SH}. The anhydrous melting temperature of coesite and the temperature of the coesite-stishovite transition at 13 GPa are taken from \cite{ZLGHF1993}. The metastable melting temperature of stishovite is taken from the melting curve of \cite{Millotetal2015}, which is consistent with the shock data of \cite{LAM1983} and static melting data of \cite{ZLGHF1993}.

The liquidus in this system provides quite severe constraints on thermodynamics in the anhydrous system. Given that SiO$_2$-H$_2$O liquids approach ideality with increasing pressure and temperature \citep{HM2012}, we feel justified in using the ideal mixing model of \cite{SS1985}. Using this model constrains the entropy of melting of stishovite to be between about 8 and 19 J/K/mol, not significantly higher than room pressure values \citep{HM2012,ZLGHF1993}. A value of 13.5 J/K/mol with G$_{OH}$=35000-20T J/mol yields the model shown in Figure  \ref{fig:SH}. Unfortunately, these entropy values of are much higher than the values estimated from the melting curve of SiO$_2$. The melting curve of coesite has a negligable dT/dP at 13 GPa \citep{ZLGHF1993}, constraining the $\Delta$V of melting to be approximately zero. Using the volume of coesite at the coesite-stishovite-liquid triple junction as a proxy for the volume of the melt, published melting curves of stishovite \citep{Millotetal2015, ZLGHF1993} imply an entropy of melting of $\sim$120 J/K/mol. There are two potential resolutions to this problem. The first is that the assumptions of the hydrous melting model are inappropriate for the high pressure SiO$_2$ system. The second is that the thermodynamic properties in the anhydrous system need to be revisited.

@@

\section{Discussion}
\subsection{Water partitioning in the mantle transition zone}
% Olivine, wadsleyite and ringwoodite

Partitioning of hydrogen between mineral and melt phases can be described using a partition coefficient D:
\begin{equation}
  D^{a/b}_{H_2O} = c^a_{H_2O} / c^b_{H_2O}
\end{equation}
where $c^i_{H_2O}$ is the concentration of H$_2$O (in wt\%) in phase $i$ \citep{KB2006}. In quenchable phases, there are a number of ways that concentration can be measured, including FTIR, SIMS and ERDA. In the case of melts, which are unquenchable, $c$ in partitioning studies has most commonly been estimated by mass balance or deficit in microprobe totals. In two studies on Fe-free wadsleyite, \cite{DDFK2005} and  \cite{LSOK2011} present estimates of water content in melts calculated from microprobe deficits. A similar study on ringwoodite was undertaken by \cite{OMY2000}. One striking observation is the extreme differences in water content estimated for melts in equilibrium with wadsleyite at 15 GPa and ringwoodite at 17-20 GPa. For example, melt in equilibrium with wadsleyite at 1400 C was estimated to contain 10.6-13.3 wt wt\% H$_2$O, while melt compositions in equilibrium with ringwoodite suggest water contents of 57-66 wt\% H$_2$O. Modelled Mg$_2$SiO$_4$ melting temperatures increase by only 40 K between 15 and 20 GPa, and entropies of melting at these two pressures are not very different (92.4 vs. 97 J/K/mol), so unless the thermodynamics of mixing in the liquid change markedly between these pressures, estimates of $c^{melt}_{H_2O}$ for the ringwoodite or wadsleyite studies (or both) must be somewhat inaccurate. We note that two of the three values provided by \cite{OMY2000} fall above the temperatures indicating that H$_2$O is the only H-bearing species in the melt \citep{SS1985}, and that the melting temperatures of wadsleyite + 20.4 wt\% H$_2$O (40 mol\% H$_2$O in Figure \ref{fig:partitioning}) in the wadsleyite studies fall below that of the ame composition at 5.5 GPa \citep{Inoue1994}. Both of these interpretations appear unlikely.

Here, we recalculate the wadsleyite/melt and ringwoodite/melt partitioning values with our melt model, assuming that the thermodynamic properties of the melt remain essentially constant between 13 and 20 GPa. Water concentrations in the solid are taken from the original studies. We restrict our analysis to data from 1100--1400 C, where the Mg:Si ratios of the melts are similar to that in forsterite. The average $D^{wad/melt}_{H_2O}$ (0.44) and $D^{rw/melt}_{H_2O}$ values (0.6) produce the lines in Figure \ref{fig:partitioning}. Note that $D^{wad/rw}_{H_2O}$ = 0.72, implying that at equilibrium, the lower transition zone should contain more water than the upper transition zone; the opposite conclusion to that reached by \cite{DDFK2005} using the same data XXXX check this!!. This prediction could explain observations of low velocity layers at ca. 520 km depth \citep{JDH2010}, in an extension of the water filter model of \cite{BK2003}.

An accurate thermodynamic model for hydrous melts can also reveal the mechanism of hydrogen incorporation into nominally anhydrous minerals. This arises from the equilibrium constant for the reaction between oxygen in the silicate, water in the melt, and OH groups in the solid: XXXX correct this!!
\begin{equation}
  K = \frac{a_{OH}^n}{a_{H_2O}^{n/2}a_O^{n/2}}
\end{equation}

Isolated protons (n=0.5), proton pairs (n=1) and proton . Making the approximation that $a_O = 1$, which is probably reasonable at low H concentrations in the solid 
Making the assumption that concentration of H$_2$O in the solid is proportional to the activity of OH in the melt, we therefore have
\begin{equation}
  c \propto a_{H_2O}^n
\end{equation}
where $n$ relates to the mechanisms of hydrogen incorporation into the solid \citep{KB2006}

@@

>>>>

\title{Hydrous melting and partitioning in and above the mantle transition zone: insights from water-rich MgO-SiO$_2$-H$_2$O experiments}
%\tnotetext[mytitlenote]{Fully documented templates are available in the elsarticle package on \href{http://www.ctan.org/tex-archive/macros/latex/contrib/elsarticle}{CTAN}.}

@@

In this study, we conducted high pressure multianvil experiments at 13 GPa between 1200 and 1900 $^{\circ}$C to investigate the liquidus in the system MgO-SiO$_2$-H$_2$O. Water-rich starting compositions were obtained by using platinic acid (H$_2$Pt(OH)$_6$) as a novel water source. As MgO:SiO$_2$ ratios decrease, the liquidus curve develops an increasingly pronounced concave-down topology. At slab temperatures, the forsterite- and enstatite-water liquidii have similar water contents. Preliminary experiments on the SiO$_2$-H$_2$O binary reveal a pronounced melting point depression up to about 20 mol\% H$_2$O. This melting point depression exceeds estimates from published models, and requires either significant non-ideality between species in the melt, or distinct differences in the ease of protonation of bridging oxygens. We discuss the implications of this work on the migration of deep-seated slab fluids, and the partitioning of water between nominally anhydrous phases and melts.
\end{abstract}

@@

\section{Introduction}
Hydrous melts have a remarkable influence on Earth's surface and interior. Melting at subduction zones is primarily driven by the release of hydrous fluids from the warming slab at about 100 km depth, producing the world's arc volcanoes. The water cycle does not stop here, however; hydrous minerals within the crust should allow some water to be transported to 250 km \citep{PS2002}, and peridotite hydration should allow water to penetrate into the mantle transition zone or deeper, providing it remains low enough in temperature \citep{KOM2005}. This deep water cycle is probably the source for some kimberlitic eruptions, as implied by the discovery of water-saturated ringwoodite in a diamond from Juina, Brazil \citep{Pearsonetal2014}. The ringwoodite stability field requires that the kimberlite must have been sourced from at least ca. 500 km depth. In addition to kimberlites and arc volcanoes, hydrous melts are known to metasomatise and refertilise the lithospheric mantle, creating the so-called MARID (mica, amphibole, rutile, ilmenite, diopside) assemblage. It has also been argued that neutral or negatively buoyant water-rich melts are to blame for low velocity and high conductivity layers observed at the 410 km discontinuity, and that these small fraction melts could act to filter out incompatible elements during upwelling \citep{BK2003}.

Processes involving deep-seated hydrous melts are controlled to a great extent by melt composition. Composition is a key variable in determining melt density, viscosity and solid-solid-melt dihedral angles, which together with the temperature and equivalent properties of the solid control the time- and lengthscales on which melt can separate from the melt. They also govern the extent to which this melt can become channelised, reducing interaction with its surroundings and increasing ascent rates. Melt compositions are also required to understand the stability of dense hydrous magnesium silicate (DHMS) phases and the water capacity of nominally anhydrous minerals (NAMS) at high pressure. Together, these control water transport and storage in the deep mantle, and as such are the primary inputs to models of the water budget of the Earth. Hydrous fluids at elevated pressures and temperatures cannot be treated as pure H$_2$O. As pressure increases, immiscibility between fluid and melt decreases, and the thermal stability of hydrous phases increases. As a result, fluids/melts released by the breakdown of hydrous phases have H$_2$O activities significantly lower than one. Understanding and predicting the \emph{P-T} conditions of DHMS phase breakdown and the water storage capacity at of NAMS therefore requires good activity-composition models for hydrous melts.

@@

In this study, we use high purity platinic acid, hexahydroxyplatinate(IV) (H$_2$Pt(OH)$_6$), as a novel water source. Platinic acid is a pale yellow compound which is stable to about 130 $^{\circ}$C \citep{Nagano2002}. It breaks down at higher temperatures, releasing four molecules of H$_2$O and forming platinum (IV) oxide PtO$_2$. Its relatively high stability, high water content and the inert nature of the breakdown product (in a system where redox reactions are absent) make it an excellent water source for high pressure hydrous melting experiments.

\section{Experimental and analytical methods}
Starting compositions were created from a mixture of high-purity brucite, quartz and platinic acid (H$_2$Pt(OH)$_6$). Before weighing, quartz was dried at 1000 $^{\circ}$C for 12 hours, while brucite was heated to 250 $^{\circ}$C, also over 12 hours. Both powders were stored in a desiccating oven at 130 $^{\circ}$C. Platinic acid is hygroscopic, so was stored in a vacuum desiccator. Powders were weighed and ground dry in an agate mortar for 30 minutes, using a mask, goggles and gloves to avoid physical contact with the platinic acid. Starting compositions were stored in glass vials in a vacuum desiccator. Compositions are listed in Table \ref{table:compositions}.

@@

\section{Modelling water solubility in melts}

@@

In this study, we use a modified version of the molecular dynamics-derived anhydrous melt model of \cite{DKS2013} using melt endmembers MgO and SiO$_2$. For MgO melt, only the reference Helmholtz free energy needed to be adjusted. We note that the melting point of MgO at high pressure is rather uncertain. In the present model \citep{DKS2013}, MgO melts at $\sim$ 4180 K at 13 GPa, similar to that determined by previous ab-initio studies \citep{Alfe2005}. Experimental results vary by about 1000 K about this value \citep[3100 -- 5373 K at 13 GPa][]{ZB1994,ZF2008}. We note that if the present model is incorrect, it is probably the calculated entropy of melting which is at fault. A higher melting temperature would therefore imply a lower entropy of melting, such that the values in $G_{\textit{melt}}$ - $G_{\textit{solid}}$ at low temperatures will have smaller errors, than would otherwise be expected from such large melting point uncertainties. For the SiO$_2$ melt, the bulk modulus of the SiO$_2$ endmember was increased to satisfy the experimental constraint that coexisting melt and coesite have similar volumes at 13--14 GPa \citep{ZLGHF1993}. The volume at 1 bar is fit to experimental data. The entropy of melting at the coesite-stishovite-melt triple point ($\sim$40 J/K/mol) was kept the same. This value is identical to the value derived from the Clapeyron slope of the stishovite liquidus \cite{SL1995}. We note that this slope is much steeper than that recently proposed by \cite{Millotetal2015}. This difference is not the result of incompatibility between experimental results at different pressures \citep{SL1995, LAM1983, Millotetal2015}, but instead the result of using a 2-parameter Simon's equation to fit the rather abrupt change in dT/dP implied by the experiments at 20--40 GPa.

A subregular solution model was adopted to describe the free energy between the MgO and SiO$_2$ melts. The two interaction parameters W$_{MS}$ and W$_{SM}$ at 14 GPa were derived from the melt composition at the forsterite (fo)-HP clinoenstatite (hen) eutectic \citep{PWMW1998} and melting temperatures at the fo, hen and fo-hen cotectic compositions. The temperature dependence around 14 GPa was calculated by assuming an ideal excess entropy contribution, and then fitting to the known melting curves of forsterite and enstatite. The pressure dependencies were fit to the melting curves. It should be noted that although the excess volumes are similar to the results of \cite{DKS2013}, the excess entropies are some 50\% lower. However, they provide an excellent fit to the forsterite and orthoenstatite melting curves (see Supplementary Materials).

The calculations in the work were undertaken with the mineral physics toolkit Burnman \citep[available from \url{https://geodynamics.org/cig/software/burnman/}][]{CHRU2014}.

\clearpage
\section{Results}
\subsection{Ex-situ observations}
Experimental details and results are shown in Table \ref{table:experiments}. Superliquidus runs revealed networks of dendritic crystals of brucite, forsterite, enstatite and stishovite. Below the liquidus, solid phases tended to aggregate at the cold end of the capsules. Periclase formed equant crystals often associated with quench overgrowths which appear brighter in BSE images. Forsterite formed tabular crystals about 100 $\mu$m long. Enstatite crystals were much smaller, forming ca. 5 $\mu$m equant grains. Stishovite also formed equant to blocky grains, but crystals only separated from the melts in chambers with a high water content. In addition to the difference in texture and shape between quench crystals liquidus phases, subliquidus phases typically contained many small inclusions of Pt-Rh oxides. A typical set of run products is shown in Figure \ref{fig:sem}.

\subsection{Subsystems}
\subsubsection{MgO-H$_2$O}
The dehydration of brucite at high pressure was previously studied by \cite{FIYKFO2005}. They reported that the stability of brucite reaches a maximum at about 9--10 GPa and 1200 $^{\circ}$C, decomposing at 1100--1150 $^{\circ}$C at 13 GPa. We observed large brucite crystals at 1200 $^{\circ}$C and 13 GPa; the temperature difference is probably within experimental uncertainties. In the present study, the composition of the fluid coexisting with brucite and periclase is constrained to be 64--66 mol\% H$_2$O. The MgO content of the fluid increases gradually with increasing pressure, as a result of the extremely high melting temperature of periclase.

Using the ideal OH$^-$ mixing model of \cite{SS1985} with no compositional dependence on the formation of OH groups, $G^{\circ}= -75000$ provides a good fit to the data if additional asymmetry is added via a binary subregular model (W$_{MgO-H_2O}$=55000 J/mol, W$_{H_2O-MgO}$=0 J/mol). This asymmetry is required to fit the activity of water at the periclase-brucite-liquid triple point to the solid thermodynamic models of \citep{HP2011} and the equation of state of water formulated by \cite{PS1995} ($0.52$). This model also has the effect of increasing the tendency for melt-fluid immiscibility at the water-rich end of the system. The second critical endpoint of the MgO-H$_2$O is probably around 13 GPa \citep{MSUP2007}, so large positive interaction enthalpies are justified in this region of the phase diagram.

@@

\clearpage
\subsubsection{SiO$_2$-H$_2$O}
The 13 GPa SiO$_2$-H$_2$O binary phase diagram is shown in Figure \ref{fig:SH}.

@@

\section{Discussion}
\subsection{Melting in the MgO-SiO$_2$-H$_2$O system}
\subsection{Water partitioning in the mantle transition zone}
% Olivine, wadsleyite and ringwoodite

Partitioning of hydrogen between mineral and melt phases can be described using a partition coefficient D:
\begin{equation}
  D^{a/b}_{H_2O} = c^a_{H_2O} / c^b_{H_2O}
\end{equation}
where $c^i_{H_2O}$ is the concentration of H$_2$O (in wt\%) in phase $i$ \citep{KB2006}. In quenchable phases, there are a number of ways that concentration can be measured, including FTIR, SIMS and ERDA. In the case of melts, which are unquenchable, $c^{melt}_{H_2O}$ in partitioning studies has most commonly been estimated by mass balance or by assuming that the deficit in microprobe totals is due to hydrogen. In two studies on Fe-free wadsleyite, \cite{DDFK2005} and \cite{LSOK2011} present estimates of water content in melts calculated from microprobe deficits. A similar study on ringwoodite was undertaken by \cite{OMY2000}. The T-X$_{H_2O}$ dependencies of the melts in equilibrium with wadsleyite or ringwoodite are strikingly different (Figure \ref{fig:partitioning}). For example, melt in equilibrium with wadsleyite at 15 GPa and 1400 $^{\circ}$C was estimated to contain 10.6-13.3 wt\% H$_2$O, while melt compositions in equilibrium with ringwoodite at 20 GPa and 1400 $^{\circ}$C suggest water contents of 57-66 wt\% H$_2$O. Modelled Mg$_2$SiO$_4$ melting temperatures increase by only 40 K between 15 and 20 GPa, and entropies of melting at these two pressures are not very different (92.4 vs. 97 J/K/mol), so unless the thermodynamics of mixing in the liquid change markedly between these pressures, estimates of $c^{melt}_{H_2O}$ for the ringwoodite or wadsleyite studies (or both) must be somewhat inaccurate. We note that two of the three values provided by \cite{OMY2000} fall above the temperatures indicating that H$_2$O is the only H-bearing species in the melt \citep{SS1985}, and that the melting temperatures of wadsleyite + 20.4 wt\% H$_2$O (X$_{H_2O}$=0.4 in Figure \ref{fig:partitioning}) in the wadsleyite studies fall below that of the same composition at 5.5 GPa \citep{Inoue1994}. Both of these interpretations appear unlikely.

Here, we recalculate the wadsleyite/melt and ringwoodite/melt partitioning values with our melt model, assuming that the thermodynamic properties of the melt remain essentially constant between 13 and 20 GPa. Water concentrations in the solid are taken from the original studies. We restrict our analysis to data from 1100--1400 $^{\circ}$C, where the Mg:Si ratios of the melts are similar to that in forsterite. At 15 GPa, $D^{wad/melt}_{H_2O}$ increases from 0.30 to 0.51 between 900 and 1200 $^{\circ}$C, and then decreases back to 0.32 at 1400 $^{\circ}$C. The three ringwoodite H$_2$O concentrations (20-23 GPa) indicate a decrease in $D^{rw/melt}_{H_2O}$ from 0.68 to 0.57 between 1300 and ca. 1400 $^{\circ}$C.

It is also possible to create a thermodynamic model for incorporation of H$_2$O into wadsleyite and ringwoodite, given some knowledge of the substitution mechanism. Hydrogen in wadsleyite and ringwoodite is incorporated as proton pairs substituting for Mg (Mg$^{2+}$ $\leftarrow$ 2H$^{+}$) \citep{Smyth1987, SHFJLM2003}. Ringwoodite potentially also exhibits a substitution involving Si (Si$^{4+}$ $\leftarrow$ Mg$^{2+}$ + 2H$^{+}$) \citep{KKMO2000}. Both of these reactions may be written in the form 
\begin{equation}
H_2O + O = (OH)_2
\end{equation}
which leads to the equilibrium relation 
\begin{eqnarray}
\Delta G = -R T \ln \left( \frac{a_{(OH)_2}}{a_{H_2O}a_{O}} \right), \\
= \Delta H - (T-T_0) \Delta S + (P-P_0) \Delta V
\end{eqnarray}
where $\Delta G$ is the Gibbs energy associated with the reaction. The second equation is a linearised form of the expression describing the change in free energy with pressure and temperature. The first equation can be simplified by making the approximation that $a_O$ is constant and that the weight percentage of H$_2$O in the solid $c_{H_2O}$ is proportional to the activity of (OH)$_2$. Both of these assumptions are probably reasonable at low values of $c_{H_2O}$. Rearranging now produces the expression:
\begin{equation}
c_{H_2O} \propto a_{H_2O} \exp{ \left( -\frac{\Delta H - (T-T_0) \Delta S + (P-P_0) \Delta V}{R T} \right) }
\end{equation}
Further simplifications can be made by taking $T_0 = 0$ K, by realising that $P_0$ can be neglected at high pressures, and by incorporating the factor $\Delta S/R$ into the exponential prefactor. Finally, it is recognised that the pressure and temperature dependence of the Gibbs free energy of water $G_{H_2O}$ is highly nonlinear. For this reason, we split the Gibbs free energy change of the anhydrous and hydrous solid from that of the liquid, yielding the equation \citep{KB2006}:
\begin{eqnarray}
c_{H_2O} = A a_{H_2O} f_{H_2O} \exp{ \left( -\frac{\Delta H + P \Delta V^{solid}}{R T} \right) }, \\
f_{H_2O} = \exp \left(\frac{\Delta G_{H_2O}}{RT} \right)
\end{eqnarray}

We calculate the fugacity of pure water $f_{H_2O}$ from the equation of state of \citep{PS1995}. We can now fit the values of A, $\Delta H$ and $\Delta V$ to the experimental data on wadsleyite and ringwoodite using our melt model. The pressure data of \cite{DDFK2005} at 1200 $^{\circ}$C and 14--17 GPa are consistent with a negligable pressure dependence on water content of wadsleyite, implying that $\Delta V \sim 10^{-5}$ m$^3$/mol, similar or slightly lower than estimates for olivine \citep[1.00 -- 1.06 $\cdot 10^{-5}$ m$^3$/mol][]{KKR1996, ZGK2004, MDAR2006}. A single data point at 18 GPa has much lower water contents, which would require $\Delta V \sim 1.25 \cdot 10^{-5}$ m$^3$/mol and would produce an almost linear curve which produces large misfits at 16 and 17 GPa. We therefore favour the lower estimate of $\Delta V$, and assume the same value for ringwoodite. Other estimated values are shown in Table \ref{table:partitioning}.

\begin{table}[]
\centering
\caption{Thermodynamic models for water solubility in wadsleyite and ringwoodite}
\label{table:partitioning}
\begin{tabular}{lllll}
Phase & $P_{\textit{\tiny{ref}}}$ (Pa) & A [kg/(kgPa)] & $\Delta H_{P_{\textit{\tiny{ref}}}}$ [J/mol] & $\Delta V$ [m$^3$/mol] \\
\hline
Wadsleyite & 1.5e10 & 4.06e-12 & 1.66e5 & 1e-5 \\
Ringwoodite & 2.0e10 & 2.19e-12 & 2.04e5 & 1e-5 (fixed) \\
\end{tabular}
\end{table}

These models allow us to predict the partition coefficient of water between wadsleyite and ringwoodite at the 520 km discontinuity ($\sim$18 GPa). The coefficent increases gradually with temperature between 0.6 at 1000 $^{\circ}$C to almost 1.0 at 2000 $^{\circ}$C. One implication is that repartitioning of water during secular cooling of the Earth is likely to be much more minor (and in the opposite direction) to that proposed by \cite{DDFK2005}. The second implication is that water saturated peridotite in the lower mantle will partially melt during upwelling, which could explain observations of low velocity layers at ca. 520 km depth \citep[e.g.][]{JDH2010}.

@@

\section{Conclusions}
